---
title: A quick look at the CCR5 data
author: Matthew Stephens and Peter Carbonetto
output: workflowr::wflow_html
---

This is for Problem Set 7, Part B. See also [Novembre et al
2005](https://doi.org/10.1371/journal.pbio.0030339).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

Load some packages used in the analysis below:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(geosphere)
library(ggplot2)
library(cowplot)
library(maps)
library(mvtnorm)
```

Load the CCR5 data:

```{r load-data}
ccr5 <- read.table("../data/CCR5/CCR5.freq.txt",header = TRUE)
```

Plot the distribution of allele frequencies:

```{r hist-allele-freq, fig.height=3.25, fig.width=3}
hist(ccr5$Freq,breaks = 16)
```

Change longitudes greater than 180 to negative numbers:

```{r fix-longitudes}
ccr5$Long <- ifelse(ccr5$Long > 180,ccr5$Long - 360,ccr5$Long)
```

We will use this function to plot the allele frequencies on a map of
Europe (and parts of Asia):

```{r draw-map}
draw_map <- function (dat)
  ggplot(dat,aes(x = Long,y = Lat)) +
    geom_path(data = map_data("world"),
              aes(x = long,y = lat,group = group),
              color = "gray",linewidth = 0.35) +
    xlim(c(-20,80)) +
    ylim(c(30,70)) +
    theme_cowplot(font_size = 12)
```

This plot shows the allele frequency data and the geographical
locations where these data were obtained. Note that a few data points
are not shown in this map.

```{r plot-data, fig.height=3.5, fig.width=5.5, warning=FALSE}
draw_map(ccr5) +
  geom_point(shape = 21,color = "black",size = 2.5,
             mapping = aes(fill = Freq)) +
  scale_fill_gradient2(low = "deepskyblue",mid = "gold",high = "red",
                       midpoint = 0.1) +
  ggtitle("data")
```

Compute pairwise distances
--------------------------

Compute distances (in km) between the points.  Dividing by 1,000 gives
distance in km.

```{r distm, fig.height=3.25, fig.width=3}
D <- distm(ccr5[1:2])/1000
hist(D,breaks = 16,xlab = "distance (km)")
```

These distances seem to broadly make sense. For example, the driving
distance from Paris to Moscow is about 3,000 km.

Deal with zero counts
---------------------

Some of the frequency estimates are zero. I deal with this by first
working out the original counts (2 * frequency * samplesize), and then
adding a pseudocount to each allele before recomputing frequencies.
The resulting column "fhat" can be transformed by log(fhat/(1-fhat))
to be something that is not entirely non-normal.

```{r compute-Z, fig.height=3.5, fig.width=3}
logit <- function (x) log(x/(1-x))
ccr5 <- transform(ccr5,count = round(2*Freq*SampleSize))
ccr5 <- transform(ccr5,fhat = (count + 1)/(2*SampleSize + 2))
ccr5 <- transform(ccr5,Z = logit(fhat))
hist(ccr5$Z,breaks = 16,xlab = "Z")
```
