---
title: "quick look at CCR5 data"
author: "Matthew Stephens"
date: 2016-02-25
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```


## Load Data

The following reads in the data, and converts longitudes >180 to corresponding
negative longitudes (this avoids warnings in geosphere package later)
```{r}
ccr5 = read.table("../data/CCR5/CCR5.freq.txt",header=TRUE)
ccr5[,1] = ifelse(ccr5[,1]>180,ccr5[,1]-360,ccr5[,1]) # changes longitudes>180 to negative
```

# Compute pairwise distances

This can be done using the geosphere package. Dividing by 1000 gives distance in km.
```{r}
geo.dist = geosphere::distm(ccr5[,1:2])/1000
hist(geo.dist)
```

# Deal with 0 counts

Some of the frequency estimates are 0. I deal with this
by first working out the original counts (frequency * samplesize *2),
and then adding a pseudocount to each allele before recomputing frequencies.
The resulting column "fhat" can be transformed by log(fhat/(1-fhat))
to be something that is not entirely non-normal...

```{r}
ccr5$count = round(ccr5$Freq* ccr5$SampleSize * 2)
ccr5$fhat = (ccr5$count+1)/(ccr5$SampleSize*2+2)
ccr5$Z = log(ccr5$fhat/(1-ccr5$fhat))
hist(ccr5$Z)
```

# Similarity of allele frequency with distance?

Here we take a quick look to check that Z has some kind of spatial trend.

```{r}
plot(x=ccr5$Lat,y=ccr5$Long,type="n")
text(round(ccr5$Z),x=ccr5$Lat,y=ccr5$Long)
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
